What does the crowd believe? A hierarchical approach to estimating subjective beliefs from empirical data

### 2016-01-31 text/01_CogSci_abstract/paper.tex
minor edits model section

We elicited prior beliefs from each participant on eight different items (listed in \tableref{tab:Items}) using three different dependent measures, which we call \emph{give-a-number}, \emph{binned histograms}, and \emph{paired comparisons}. Each participant rated each item using each dependent measure. Trial order was randomized.

\begin{table*}
  \footnotesize
  \centering
  \begin{tabular}{l p{2.2cm} p{2.5cm} p{3.5cm} p{3cm} p{2.5cm}}
	\toprule
	Item & Bins (units, min, max, step) & Context sentence & GAN question & BH frame & PC frame \\
	\midrule
    coffee & degrees, $<$44, $>200$, 13& X has just fetched himself a cup of coffee from the office vending machine. & What do you think the temperature of his coffee is?  & his coffee was the following temperatures & The temperature of his coffee is N degrees.\\
	commute & minutes, 0, $>$98, 7 & X commuted to work yesterday. & How many minutes do you think she spent commuting yesterday? & she commuted for the following numbers of minutes yesterday & She spent N minutes commuting.\\
	joke & children, 0, 14, 1  & X told a joke to 14 kids. & How many of the kids do you think laughed? & the following number of kids laughed & N of the children laughed.\\

	laptop & dollars, 0, $>$7500, 500& X bought a laptop. & How much do you think it cost? & the laptop cost the following numbers of dollars & The laptop cost \$N.\\
    marbles & marbles, 0, 14, 1 & X threw 14 marbles into a pool. & How many of the marbles do you think sank? & the following number of marbles sank & N of the marbles sank.\\

	movies & minutes, 0, $>210$, 16& X just went to the movies to see a blockbuster. & How many minutes long do you think the movie was? & the movie was the following numbers of minutes long & The movie was N minutes long.\\

	TV & hours, 0, $>$43, 3& X watched TV last week. & How many hours do you think he spent watching TV last week? & he watched TV for the following numbers of hours last week & He spent N hours watching TV.\\

	watch & dollars, 0, $>$750, 50 & X bought a watch.& How much do you think it cost? & it cost the following numbers of dollars & The watch cost \$N.\\
    \bottomrule
  \end{tabular}
  \caption{Experimental items. \emph{X} was a randomly generated name (different on each trial). \emph{N} was one of the bins.}
  \label{tab:Items}
\end{table*}

On  \emph{give-a-number} (GAN) trials, participants saw the context sentence and GAN question for each item (see \tableref{tab:Items}) and were asked to provide one number by adjusting a slider with endpoints labeled as the lowest and highest number for that item. Min and max numbers are shown in \tableref{tab:Items} and were taken from previous studies that had used these items \cite{DegenTessler2015:Wonky-worlds:-L,SchollerFranke2015:Semantic-values,KaoWu2014:Nonliteral-Unde}\jd{any others?}. The selected number appeared above the slider so participants knew exactly what the value of the slider would be.

On \emph{binned histogram} (BH) trials, participants saw an item's context sentence and were asked to \emph{Please rate how likely it is that Y} (where Y corresponded to  the appropriate BH frame in \tableref{tab:Items}). They then adjusted 15 sliders, one per bin, with five points labeled \emph{impossible}, \emph{not very likely}, \emph{neutral}, \emph{very likely}, and \emph{extremely likely}. Bins were determined by dividing up the entire interval spanned by the item's minimum and maximum value into equally sized bins (see \tableref{tab:Items} for each item's bin size).

On \emph{paired comparison} (PC) trials, participants  saw an item's context sentence and were asked to click on one of two options which were shown side by side -- whichever one they thought was more likely.  Each option used the appropriate PC frame in\tableref{tab:Items} and numbers were filled in by comparing the following bins: 1 vs 2, 2 vs 6, 6 vs 11, 11 vs 14, and 14 vs 15. Paired comparison trials always occurred as a block of five trials, with order of comparisons randomized within block.

Results are shown in \figref{fig:Results}. Data analysis and modeling is reported in the next section.\jd{do we need to say anything about this or just go straight to the model?}

\begin{itemize}
\item why within-participant? \jd{because doing it between participants would mean not knowing if any differences in results are due to differences in beliefs or differences in the measure. ie we didn't want to confound participants and measures. do we need to talk about this?}
%\item what was the exact randomization / procedure?
%\item items are given in Table~\ref{tab:Items}
%\item include bins as well?

%\item results are in Figure~\ref{fig:Results}
\end{itemize}

@@

All three pieces of data are to be explained as functions of subjective beliefs $P_{ij}$, where
$P_{ij}$ is a probability vector of length 15 with $P_{ijk}$ being participant $i$'s belief about
the relative likelihood of bin $k$ for item $j$. Each $P_{ij}$ defines a likelihood for our
data, via appropriate link functions (to be spelled out presently). Variance in subjective
beliefs is harnessed by a population-level hyper-prior with central tendency $Q_{j}$, i.e., a
stochastic vector of length 15, which we call a ``population-level belief'' about item $j$. The
structure of this model is pictured in \figref{fig:ModelGraph}.

@@

Finally, consider the link function for bin comparisons. We are interested in the likelihood with which
participant $i$ selects the higher bin for item $j$ in bin comparison condition $l$. Suppose $l$ is
about comparing the lower bin $b_l$ to the higher $b_h$. The perhaps most natural approach
would be to link the likelihood of choosing $b_h$ over $b_l$ to the difference between
$P_{ijb_h}$ and $P_{ijb_l}$. This, however, does not appear to be what participants were doing. For
example, take the ``marbles'' case where we see almost unanimous choices of bin 11 over bin 6,
despite seeing no pronounced difference in slider ratings for these bins. Indeed, a model that
implements this idea failed to capture the relevant regularities in the data blatantly.

Another plausible link function that accommodates for this is to assume that what matters in
bin comparisons is the distance to the mode of $P_{ij}$: soft-max prefer the bin that is closer
to the mode of $P_{ij}$; select randomly if both bins are equally far from the
prototype.\footnote{Notice that
  $(1 + \exp(2b(1-p^{\text{high}}_{ijl})) )^{-1} = \frac{\exp(b p^{\text{high}})}{\exp(b
    p^{\text{high}}) + \exp(b p^{\text{low}})}$
  with $p^\text{low}_{ijl} = 2 - p^\text{high}_{ijl}$.}
\begin{align*}
  c_{ijl} & \sim \text{Bern}( (1 + \exp(2b(1-p^{\text{high}}_{ijl})) )^{-1} ) \ \ \ 
  b  \sim \text{Gamma}(2,1) \\
  p^\text{high}_{ijl} & = \begin{cases}
    2 & \text{if $mode(P_{ij})$ is closer to higher} \\
    & \text{ bin of $l$ than to lower bin } \\ 1 & \text{if equal
      distance} \\ 0 & \text{otherwise}  
  \end{cases}
\end{align*}

\ndg{wow, this is weird! you tried softmax ratio of bin probs?}

\jd{hm, doesn't seem too weird to me!}

@@

@@

\ndg{it'd be nice to see some posteriors for some of the (more meaningful) parameters. eg w, kappa, ...}

\ndg{what about the success of the other DMs? looks like give-n is not so hot. worth saying that if so.}

>>>>

We elicited prior beliefs for eight different items (listed in
\tableref{tab:Items}) using three different dependent measures: \emph{binned histograms},
\emph{give-a-number} and \emph{paired comparisons}. Each participant rated each item using each
dependent measure. Trial order was randomized.

\begin{table*}
  \footnotesize
  \centering
  \begin{tabular}{l p{2.2cm} p{2.5cm} p{3.5cm} p{3cm} p{2.5cm}}
	\toprule
	Item & Bins (units, min, max, step) & Context sentence & GAN question & BH frame & PC frame \\
	\midrule
    coffee & degrees, $<$44, $>200$, 12& X has just fetched himself a cup of coffee from the office vending machine. & What do you think the temperature of his coffee is?  & His coffee was the following temperatures & The temperature of his coffee is N degrees.\\
	commute & minutes, 0, $>$98, 7 & X commuted to work yesterday. & How many minutes do you think she spent commuting yesterday? & She commuted for the following numbers of minutes yesterday & She spent N minutes commuting.\\
	joke & children, 0, 14, 1  & X told a joke to 14 kids. & How many of the kids do you think laughed? & the following number of kids laughed & N of the children laughed.\\

	laptop & dollars, 0, $>$7500, 500& X bought a laptop. & How much do you think it cost? & The laptop cost the following numbers of dollars & The laptop cost \$N.\\
    marbles & marbles, 0, 14, 1 & X threw 14 marbles into a pool. & How many of the marbles do you think sank? & The following number of marbles sank & N of the marbles sank.\\

	movies & minutes, 0, $>210$, 16& X just went to the movies to see a blockbuster. & How many minutes long do you think the movie was? & The movie was the following numbers of minutes long & The movie was N minutes long.\\

	TV & hours, 0, $>$43, 3& X watched TV last week. & How many hours do you think he spent watching TV last week? & He watched TV for the following numbers of hours last week & He spent N hours watching TV.\\

	watch & dollars, 0, $>$750, 50 & X bought a watch.& How much do you think it cost? & It cost the following numbers of dollars & The watch cost \$N.\\
    \bottomrule
  \end{tabular}
  \caption{Experimental items. \emph{X} was a randomly generated name (different on each trial). \emph{N} was one of the bins.}
  \label{tab:Items}
\end{table*}

On \emph{give-a-number} (GAN) trials, participants saw the context sentence and GAN question
for each item (see \tableref{tab:Items}) and were asked to provide one number by adjusting a
slider with endpoints labeled as the lowest and highest number for that item. Min and max
numbers are shown in \tableref{tab:Items} and were taken from previous studies that had used
these items
\cite{DegenTessler2015:Wonky-worlds:-L,SchollerFranke2015:Semantic-values,KaoWu2014:Nonliteral-Unde}. The
selected number appeared above the slider so participants knew exactly what the value of the
slider would be.

On \emph{binned histogram} (BH) trials, participants saw an item's context sentence and were
asked to \emph{Please rate how likely it is that Y} (where Y corresponded to the appropriate BH
frame in \tableref{tab:Items}). They then adjusted 15 sliders, one per bin, with five points
labeled \emph{impossible}, \emph{not very likely}, \emph{neutral}, \emph{very likely}, and
\emph{extremely likely}. Bins were determined by dividing up the entire interval spanned by the
item's minimum and maximum value into equally sized bins (\tableref{tab:Items}).

On \emph{paired comparison} (PC) trials, participants saw an item's context sentence and had to
click on one of two options shown side by side -- whichever one they thought was more likely.
Each option used the appropriate PC frame in\tableref{tab:Items} and numbers were filled in by
comparing the following bins: 1 vs 2, 2 vs 6, 6 vs 11, 11 vs 14, and 14 vs 15. Paired
comparison trials always occurred as a block of five trials, with order of comparisons
randomized within block.

Results are shown in \figref{fig:Results}. \figref{fig:Results} shows mean slider ratings in
relation to frequencies of number choices in the \emph{give-a-number} task. It looks like the
latter could be samples from the former, with a tendency to modal choices. But, despite the use
of slider, some \emph{give-a-number} results still seem to reflect a tendency towards round or
salient numbers. E.g., in item \emph{coffee} more than 40\% of participants guessed that a
coffee from the vending machine was ca. 150 degrees Fahrenheit (bin 10). The results from the
\emph{paired comparison} seem consistent with the mean slider ratings as well, albeit not as
straightforwardly as one might think. This is clearest from the \emph{marbles} case. While the
mean slider ratings indicate that bins 2, 6 and 11 were all felt to be equally likely on
average, almost every participant chose the higher bin in direct bin comparisons 2 vs 6 and 6
vs 11. A potential explanation for this, which we will implement formally in the
data-generating model, is that the \emph{paired comparison} condition invited subjects to think
about what they thought was mostly likely the case (usually: bin 15 in the
\emph{marbles} case) and to choose the bin that is closest to that.

@@

All three pieces of data are to be explained as functions of subjective beliefs $P_{ij}$, with
$P_{ijk}$ being participant $i$'s belief about the relative likelihood of bin $k$ for item
$j$. Each $P_{ij}$ defines a likelihood for our data, via appropriate link functions (to be
spelled out presently). Variance in subjective beliefs is harnessed by a population-level
hyper-prior with central tendency $Q_{j}$, i.e., a stochastic vector of length 15. The
structure of this model is pictured in \figref{fig:ModelGraph}.

@@

Finally, consider the link function for bin comparisons. We are interested in the likelihood
with which participant $i$ selects the higher bin for item $j$ in bin comparison condition
$l$. Suppose $l$ is about comparing the lower bin $b_l$ to the higher $b_h$. The perhaps most
natural approach would be to link the likelihood of choosing $b_h$ over $b_l$ to the difference
between $P_{ijb_h}$ and $P_{ijb_l}$. However, as discussed above, This does not appear to be
what participants were doing. Indeed, a model that implements this idea failed to capture the
relevant regularities in the data blatantly. Another plausible link function is to assume that
what matters in bin comparisons is the distance to the mode of $P_{ij}$: soft-max prefer the
bin that is closer to the mode of $P_{ij}$; select randomly if both bins are equally far from
the prototype.\footnote{Notice that
  $(1 + \exp(2b(1-p^{\text{high}}_{ijl})) )^{-1} = \frac{\exp(b p^{\text{high}})}{\exp(b
    p^{\text{high}}) + \exp(b p^{\text{low}})}$
  with $p^\text{low}_{ijl} = 2 - p^\text{high}_{ijl}$.}
\begin{align*}
  c_{ijl} & \sim \text{Bern}( (1 + \exp(2b(1-p^{\text{high}}_{ijl})) )^{-1} ) \ \ \ 
  b  \sim \text{Gamma}(2,1) \\
  p^\text{high}_{ijl} & = \begin{cases}
    2 & \text{if $mode(P_{ij})$ is closer to higher} \\
    & \text{ bin of $l$ than to lower bin } \\ 1 & \text{if equal
      distance} \\ 0 & \text{otherwise}  
  \end{cases}
\end{align*}

@@

Means and bounds of 95\% HDIs of the posteriors for model parameters are in
Table~\ref{tab:SummaryStats}. Posterior credible levels of $w$ allow for a limited amount of
slack around $Q_j$. Values for $\kappa$ indicate that, on average, subjects may not have had a
preference for or against extreme slider ratings. Relatively high values of $a$ indicate that
subjects, on average, had a strong tendency to choose their mode in the \emph{give-a-number}
task.
\begin{table}[ht]
\centering
\begin{tabular}{rlllll}
 & $w$ & $\kappa$ & $\sigma$ & $a$ & $b$ \\ \hline
  lower & 14.65 &  0.98 &  0.26 & 22.04 &  1.11 \\ 
  mean & 15.55 &  0.99 &  0.28 & 27.43 &  1.27 \\ 
  upper & 16.48 &  0.99 &  0.31 & 32.95 &  1.42 \\ 
   \hline
\end{tabular}
\caption{Summary statistics for posterior on parameters}
\label{tab:SummaryStats}
\end{table}

@@
